---
title: "Consensus Modules"
author: "George Eusebio Kuodza"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
vignette: >
  %\VignetteIndexEntry{Consensus Modules}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE
)
```

## Introduction

In this vignette we use Comethyl to construct a **consensus** weighted
region comethylation network — a network built simultaneously across two
or more datasets (e.g., multiple timepoints or two biological groups)
that share the same genomic region definitions. Because the modules are
identified jointly, any module that emerges represents comethylation
structure that is reproducible across datasets rather than idiosyncratic
to one.

A typical use case is a longitudinal study with samples collected at
multiple timepoints (Baseline, 36–38 weeks, Postpartum), or a
sex-stratified study where female and male samples are analyzed
separately but using a common region set so that modules can be compared
directly.

The dataset used here consists of three maternal blood WGBS timepoints
from the Growing Up in Singapore Towards Healthy Outcomes (GUSTO) cohort
provided as test data in
[`lasallegrp/bgw_wgbs_comethyl`](https://github.com/lasallegrp/bgw_wgbs_comethyl/tree/main/tests/data/processed/cytosine_reports).
The three timepoints — **Baseline**, **36–38 weeks**, and **Postpartum**
— each have their own set of CpG_report files and sample metadata.

> **Relationship to other vignettes.** The CpG Cluster Analysis and Gene
> Body Analysis vignettes document the single-dataset workflow. This
> vignette covers the multi-dataset extension. Many of the underlying
> comethyl functions are the same; the key difference is the reference
> region strategy and the use of `blockwiseConsensusModules()` from WGCNA.

---

## Environment Setup

All analyses should be run within the comethyl pixi environment to
ensure reproducibility. If you have not yet set up the environment, see
the [Get Started vignette](https://lasallegrp.github.io/comethyl/articles/comethyl.html) for full installation instructions.

**Activate the environment before launching R:**

```bash
# From the repository root
pixi shell       # activates the environment
R                # launch R from within it

# Or run scripts directly
pixi run Rscript scripts/consensus/00_import_cpg_reports_consensus.R --help
```

---

## Setup

```{r libraries}
library(tidyverse)
library(comethyl)
library(WGCNA)
library(openxlsx)
```

## Set Global Options

As in the single-dataset workflow, we disable WGCNA multi-threading
when working with large region sets (> 150,000 regions). For smaller
sets or interactive exploration you can enable threads with
`WGCNA::enableWGCNAThreads()`.

```{r global-options}
options(stringsAsFactors = FALSE)
Sys.setenv(R_THREADS = 1)
WGCNA::disableWGCNAThreads()
```

---

## Overview of the Consensus Approach

The consensus workflow adds several steps before and after the steps you
would recognise from the single-dataset vignettes. The overall logic is:

1. **One reference dataset** defines the canonical CpG filter and
   genomic regions. All other datasets must use exactly these regions.
2. **Non-reference datasets** are aligned to the reference CpG universe
   and their region methylation is calculated over the reference regions.
3. After all datasets have region methylation matrices, a
   **shared region filter** (Script 05b) keeps only regions that are
   present and sufficiently variable across every dataset.
4. PCs are derived and methylation is adjusted independently for each
   dataset (Scripts 06–07).
5. A **single consensus soft-power** is chosen that satisfies scale-free
   topology criteria in every dataset simultaneously (Script 08).
6. `blockwiseConsensusModules()` identifies modules that are consistent
   across datasets (Script 09). The resulting module assignments are
   shared — a region belongs to the same module regardless of which
   dataset is examined.
7. Downstream analyses (membership, ME-trait correlations, annotation,
   enrichment) use these shared assignments.

### Choosing a Reference Dataset

The reference dataset determines which CpGs are retained and which
genomic regions are defined. This choice should be guided by your
research question:

- If you have a **baseline or pre-treatment timepoint**, this is usually
  the natural reference because it defines the methylation landscape
  before any intervention.
- If you are comparing **two biological groups** (e.g. female vs. male),
  choose the group with the larger sample size or the group most central
  to your hypothesis as the reference.
- If no single dataset is clearly primary, choose the one with the
  **highest sequencing depth or most samples**, as it will produce the
  most stable region definitions.

> **Practical note.** The reference only affects which CpGs and regions
> are defined in Scripts 01–02. Once regions are defined, all datasets
> are treated symmetrically in Script 05b onward.

---

## 1. Import CpG Reports (Script 00)

**Run separately for each dataset/timepoint.**

`getCpGs()` reads Bismark CpG_report files into a single BSseq object.
This step is identical to the single-dataset workflow but is run once
per dataset. The script also computes and plots CpG totals at different
coverage and per-sample cutoffs so you can assess data quality before
choosing a filter.

```{r script-00, eval=FALSE}
# Interactive equivalent of Script 00 — run for each dataset separately

# ── Baseline (reference) ─────────────────────────────────────────────────────
colData_base <- openxlsx::read.xlsx("sample_info_Baseline.xlsx", rowNames = TRUE)

setwd("cytosine_reports/Baseline/")
bs_base_unfiltered <- getCpGs(
  colData  = colData_base,
  file     = "comethyl_output/consensus/00_cpg_extraction/Baseline/Unfiltered_BSseq.rds"
)
setwd("../../")

CpGtotals_base <- getCpGtotals(
  bs_base_unfiltered,
  file = "comethyl_output/consensus/00_cpg_extraction/Baseline/CpG_Totals.txt"
)
plotCpGtotals(
  CpGtotals_base,
  file = "comethyl_output/consensus/00_cpg_extraction/Baseline/CpG_Totals.pdf"
)

# ── 36–38 weeks ──────────────────────────────────────────────────────────────
colData_36wk <- openxlsx::read.xlsx("sample_info_36-38wk.xlsx", rowNames = TRUE)

setwd("cytosine_reports/36-38wk/")
bs_36wk_unfiltered <- getCpGs(
  colData = colData_36wk,
  file    = "comethyl_output/consensus/00_cpg_extraction/36-38wk/Unfiltered_BSseq.rds"
)
setwd("../../")

# Repeat getCpGtotals / plotCpGtotals for 36-38wk ...

# ── Postpartum ────────────────────────────────────────────────────────────────
# ... same pattern for Postpartum
```

---

## 2. Filter CpGs and Define Reference Regions (Script 01)

**Run on the reference dataset only.**

We filter CpGs in the reference dataset and use `getRegions()` to
define genomic regions. These regions are the canonical definitions
reused by all other datasets.

```{r script-01}
# ── Filter reference CpGs ────────────────────────────────────────────────────
bs_base <- filterCpGs(
  bs         = bs_base_unfiltered,
  cov        = 3,
  perSample  = 0.75,
  file       = "comethyl_output/consensus/01_reference_filter_regions/Baseline/cov3_75pct/Filtered_BSseq.rds"
)

# ── Define regions ────────────────────────────────────────────────────────────
regions_raw <- getRegions(
  bs   = bs_base,
  file = "comethyl_output/consensus/01_reference_filter_regions/Baseline/cov3_75pct/Regions.txt"
)

plotRegionStats(regions_raw, maxQuantile = 0.99,
                file = "comethyl_output/consensus/01_reference_filter_regions/Baseline/cov3_75pct/Region_Plots.pdf")
plotSDstats(regions_raw, maxQuantile = 0.99,
            file = "comethyl_output/consensus/01_reference_filter_regions/Baseline/cov3_75pct/SD_Plots.pdf")

regionTotals <- getRegionTotals(
  regions_raw,
  file = "comethyl_output/consensus/01_reference_filter_regions/Baseline/cov3_75pct/Region_Totals.txt"
)
plotRegionTotals(
  regionTotals,
  file = "comethyl_output/consensus/01_reference_filter_regions/Baseline/cov3_75pct/Region_Totals.pdf"
)
```

---

## 3. Filter Reference Regions (Script 02)

**Run on the reference dataset only.**

`filterRegions()` applies a minimum coverage (`covMin`) and methylation
standard deviation (`methSD`) threshold to the raw regions. The
resulting `Filtered_Regions.txt` is the canonical region set for all
downstream steps.

### Choosing the SD threshold in the reference dataset

> **Key decision.** The `methSD` threshold in Script 02 controls how
> variable a region must be *in the reference dataset* to be retained.
>
> - If your datasets represent the same biological group at different
>   timepoints (e.g., longitudinal), a moderate threshold (0.05–0.08)
>   in the reference dataset is sensible.
> - If your datasets represent **different biological groups** (e.g.,
>   female vs. male), a region that is variable in one group may be
>   nearly invariant in the other. In this case we recommend setting
>   `methSD = 0` (or a deliberately permissive value such as 0.02) here,
>   and applying the decisive joint-variability filter in Script 05b
>   instead. This prevents the reference group from excluding regions
>   that are biologically meaningful only in the other group.

```{r script-02}
# ── Conservative/permissive reference SD filter ───────────────────────────────
# For longitudinal data: use a moderate methSD (e.g. 0.05–0.08)
# For sex-stratified or group comparisons: use methSD = 0 and rely on Script 05b

regions_filtered <- filterRegions(
  regions   = regions_raw,
  covMin    = 4,
  methSD    = 0.05,           # set to 0 for sex-stratified designs
  file      = "comethyl_output/consensus/02_reference_region_filter/Baseline/cov3_75pct/covMin4_methSD0p05/Filtered_Regions.txt"
)

plotRegionStats(regions_filtered, maxQuantile = 0.99,
                file = "comethyl_output/consensus/02_reference_region_filter/Baseline/cov3_75pct/covMin4_methSD0p05/Filtered_Region_Plots.pdf")
```

---

## 4. Build Reference Region Methylation (Script 03)

**Run on the reference dataset only.**

`getRegionMeth()` calculates region-level methylation from the filtered
BSseq object over the canonical reference regions.

```{r script-03}
meth_base <- getRegionMeth(
  regions = regions_filtered,
  bs      = bs_base,
  file    = "comethyl_output/consensus/03_region_methylation/Baseline/cov3_75pct/covMin4_methSD0p05/Region_Methylation.rds"
)
```

---

## 5. Align Other Datasets to Reference CpGs (Script 04)

**Run on non-reference datasets only.**

Non-reference datasets are aligned to the CpG universe of the reference
filtered BSseq object using `subsetByOverlaps()`. This ensures that all
datasets cover exactly the same CpGs before region methylation is
calculated.

```{r script-04}
# Load reference filtered BSseq as the CpG universe
bs_base_filtered <- readRDS(
  "comethyl_output/consensus/01_reference_filter_regions/Baseline/cov3_75pct/Filtered_BSseq.rds"
)

# ── 36–38 weeks ──────────────────────────────────────────────────────────────
setwd("cytosine_reports/36-38wk/")
bs_36wk_unfiltered <- getCpGs(
  colData = colData_36wk,
  file    = "comethyl_output/consensus/04_align_to_reference_cpgs/36-38wk/Unfiltered_BSseq.rds"
)
setwd("../../")

# Restrict to reference CpG universe
bs_36wk_aligned <- subsetByOverlaps(bs_36wk_unfiltered, bs_base_filtered)

# Apply a permissive filter (very low thresholds — we just want NA-free data)
bs_36wk_aligned <- filterCpGs(
  bs        = bs_36wk_aligned,
  cov       = 1,
  perSample = 0.01,
  file      = "comethyl_output/consensus/04_align_to_reference_cpgs/36-38wk/Filtered_BSseq.rds"
)

# Repeat for Postpartum ...
```

---

## 6. Build Non-Reference Region Methylation (Script 05)

**Run on non-reference datasets only.**

Region methylation is calculated for each non-reference dataset using
the same canonical reference regions from Script 02.

```{r script-05}
regions_filtered <- read.table(
  "comethyl_output/consensus/02_reference_region_filter/Baseline/cov3_75pct/covMin4_methSD0p05/Filtered_Regions.txt",
  header = TRUE, sep = "\t"
)

meth_36wk <- getRegionMeth(
  regions = regions_filtered,
  bs      = bs_36wk_aligned,
  file    = "comethyl_output/consensus/03_region_methylation/36-38wk/cov3_75pct/covMin4_methSD0p05/Region_Methylation.rds"
)

# Repeat for Postpartum ...
```

---

## 7. Filter Shared, Complete Regions (Script 05b)

This script is unique to the consensus workflow. It takes the region
methylation matrices from all datasets and identifies the subset of
regions that are:

- **Complete** — present and non-missing in every dataset
- **Jointly variable** — meeting an SD threshold in every dataset
  (or in a user-specified number of datasets)

### Selection modes

| Mode | Description | When to use |
|------|-------------|-------------|
| `complete_only` | Regions present and NA-free in all datasets | Legacy/simple case |
| `joint_sd_all` | Complete AND SD ≥ threshold in **every** dataset | **Recommended for sex-stratified or group comparisons** |
| `joint_sd_any` | Complete AND SD ≥ threshold in **at least one** dataset | Discovery; may retain invariant regions in some datasets |
| `joint_sd_min_n` | Complete AND SD ≥ threshold in at least N datasets | Flexible middle ground |

> **Recommended SD value.** A threshold of `joint_meth_sd = 0.07` 
> (equivalent to 7% methylation SD) works well for WGBS blood data.
> Examine the per-region SD distribution plots in Script 02 to calibrate
> this for your data. Regions below this threshold typically lack
> sufficient biological variability to contribute meaningful signal to
> the network.

```{r script-05b}
# Load all region methylation matrices
meth_base     <- readRDS("comethyl_output/consensus/03_region_methylation/Baseline/cov3_75pct/covMin4_methSD0p05/Region_Methylation.rds")
meth_36wk     <- readRDS("comethyl_output/consensus/03_region_methylation/36-38wk/cov3_75pct/covMin4_methSD0p05/Region_Methylation.rds")
meth_pp       <- readRDS("comethyl_output/consensus/03_region_methylation/Postpartum/cov3_75pct/covMin4_methSD0p05/Region_Methylation.rds")

# Find regions present in all datasets (complete_only — minimal mode)
shared_regions <- Reduce(intersect,
  list(rownames(meth_base), rownames(meth_36wk), rownames(meth_pp))
)

meth_base_shared <- meth_base[shared_regions, ]
meth_36wk_shared <- meth_36wk[shared_regions, ]
meth_pp_shared   <- meth_pp[shared_regions, ]

# ── Joint SD filter (joint_sd_all) ───────────────────────────────────────────
# Recommended when comparing groups; apply after all methylation matrices exist
joint_meth_sd <- 0.07

sd_base <- apply(meth_base_shared, 1, sd, na.rm = TRUE)
sd_36wk <- apply(meth_36wk_shared, 1, sd, na.rm = TRUE)
sd_pp   <- apply(meth_pp_shared,   1, sd, na.rm = TRUE)

pass_sd <- (sd_base >= joint_meth_sd) &
           (sd_36wk >= joint_meth_sd) &
           (sd_pp   >= joint_meth_sd)

message("Regions passing joint SD filter: ", sum(pass_sd), " / ", length(pass_sd))

meth_base_final <- meth_base_shared[pass_sd, ]
meth_36wk_final <- meth_36wk_shared[pass_sd, ]
meth_pp_final   <- meth_pp_shared[pass_sd, ]

# Save final matrices for Scripts 06 onward
saveRDS(meth_base_final,
        "comethyl_output/consensus/05b_shared_complete_regions/Baseline_Methylation_Final.rds")
saveRDS(meth_36wk_final,
        "comethyl_output/consensus/05b_shared_complete_regions/36-38wk_Methylation_Final.rds")
saveRDS(meth_pp_final,
        "comethyl_output/consensus/05b_shared_complete_regions/Postpartum_Methylation_Final.rds")
```

---

## 8. PC Diagnostics (Script 06)

**Run separately per dataset.**

`getPCs()` derives the top principal components for each dataset. The
resulting PC-trait correlation heatmaps and variance-explained plots
help you decide which PCs to include in the methylation adjustment
model. PCs associated with technical covariates (batch, sequencing
depth) should generally be included; PCs associated with biological
traits of interest should generally be excluded.

```{r script-06}
# ── Baseline ──────────────────────────────────────────────────────────────────
mod_base <- model.matrix(~1, data = pData(bs_base))
PCs_base  <- getPCs(
  meth  = meth_base_final,
  mod   = mod_base,
  file  = "comethyl_output/consensus/06_pc_diagnostics/Baseline/PCs.rds"
)

# Examine PC-trait associations to choose which PCs to regress out
# Plots are saved to the output directory specified by Script 06

# Repeat for 36-38wk and Postpartum ...
```

---

## 9. Adjust Methylation (Script 07)

**Run separately per dataset.**

`adjustRegionMeth()` regresses the selected PCs out of the region
methylation matrix. Script 07 supports three adjustment strategies:

- **v1 (all PCs):** regress out all top PCs. Simplest approach.
- **v2 (exclude protected PCs):** exclude PCs significantly correlated
  with traits you want to preserve (e.g., diagnosis, sex).
- **v3 (technical PCs only):** regress out only PCs correlated with
  known technical covariates.

Examine the sample dendrogram after adjustment — samples should no
longer cluster strongly by batch.

```{r script-07}
# ── v1: adjust for all PCs ────────────────────────────────────────────────────
methAdj_base <- adjustRegionMeth(
  meth  = meth_base_final,
  PCs   = PCs_base,
  file  = "comethyl_output/consensus/07_methylation_adjustment/Baseline/v1_all_pcs/Adjusted_Region_Methylation.rds"
)

getDendro(methAdj_base, distance = "euclidean") %>%
  plotDendro(file = "comethyl_output/consensus/07_methylation_adjustment/Baseline/v1_all_pcs/Sample_Dendrogram.pdf",
             expandY = c(0.25, 0.08))

# Repeat for 36-38wk and Postpartum ...
```

---

## 10. Select Consensus Soft Power (Script 08)

`getSoftPower()` is run on each adjusted methylation matrix, and a
single soft-thresholding power is selected that achieves scale-free
topology (R² ≥ 0.8) with a negative slope in **every** dataset. Using a
shared power ensures the network analysis is directly comparable across
datasets.

```{r script-08}
sft_base  <- getSoftPower(methAdj_base,  corType = "pearson",
                          file = "comethyl_output/consensus/08_soft_power/Baseline/SoftPower_pearson.rds")
sft_36wk  <- getSoftPower(methAdj_36wk,  corType = "pearson",
                          file = "comethyl_output/consensus/08_soft_power/36-38wk/SoftPower_pearson.rds")
sft_pp    <- getSoftPower(methAdj_pp,    corType = "pearson",
                          file = "comethyl_output/consensus/08_soft_power/Postpartum/SoftPower_pearson.rds")

plotSoftPower(sft_base, file = "comethyl_output/consensus/08_soft_power/Baseline/SoftPower_Plots.pdf")
plotSoftPower(sft_36wk, file = "comethyl_output/consensus/08_soft_power/36-38wk/SoftPower_Plots.pdf")
plotSoftPower(sft_pp,   file = "comethyl_output/consensus/08_soft_power/Postpartum/SoftPower_Plots.pdf")

# Choose the lowest power where R² >= 0.8 AND slope is negative in all datasets
chosen_power <- 12   # update based on your plots
message("Chosen consensus soft power: ", chosen_power)
```

---

## 11. Identify Consensus Modules (Script 09)

`blockwiseConsensusModules()` identifies comethylation modules that are
consistent across all datasets. Regions are first organized into blocks,
a full network analysis is performed on each block within every dataset,
and modules are merged if their eigengenes are highly correlated. The
resulting module assignments are identical for all datasets.

Key arguments to consider:

- **`power`**: the consensus soft power selected in Script 08.
- **`corType`**: `"bicor"` is more robust to outlier samples; `"pearson"` 
  has higher sensitivity.
- **`deepSplit`**: controls module granularity (0–4); higher values give 
  more, smaller modules.
- **`minModuleSize`**: minimum number of regions per module; increase to 
  avoid very small modules.
- **`mergeCutHeight`**: modules with eigengene correlation above this 
  threshold are merged; lower values produce fewer, broader modules.

```{r script-09}
# Build the multiExpr list required by blockwiseConsensusModules
multiExpr <- list(
  Baseline  = list(data = t(methAdj_base)),
  "36-38wk" = list(data = t(methAdj_36wk)),
  Postpartum = list(data = t(methAdj_pp))
)

consensus_modules <- blockwiseConsensusModules(
  multiExpr          = multiExpr,
  power              = chosen_power,
  corType            = "bicor",
  networkType        = "signed",
  TOMType            = "signed",
  deepSplit          = 4,
  minModuleSize      = 5,
  mergeCutHeight     = 0.1,
  maxBlockSize       = 40000,
  networkCalibration = "full quantile",
  saveTOMs           = TRUE,
  numericLabels      = FALSE,
  verbose            = 3
)

saveRDS(consensus_modules,
        "comethyl_output/consensus/09_consensus_modules/shared/Consensus_Modules.rds")

# Region-module assignments (shared across all datasets)
module_colors <- consensus_modules$colors
table(module_colors)
```

---

## 12. Examine Consensus Module Diagnostics (Script 10)

After module detection, we examine the consensus eigengene network to
understand relationships among modules and among samples.

```{r script-10}
MEs_base <- consensus_modules$multiMEs$Baseline$data
MEs_36wk <- consensus_modules$multiMEs[["36-38wk"]]$data
MEs_pp   <- consensus_modules$multiMEs$Postpartum$data

# ── Module correlation structure ──────────────────────────────────────────────
moduleDendro_base <- getDendro(MEs_base, distance = "bicor")
plotDendro(moduleDendro_base, labelSize = 4, nBreaks = 5,
           file = "comethyl_output/consensus/10_diagnostics/Baseline/Module_ME_Dendrogram.pdf")

moduleCor_base <- getCor(MEs_base, corType = "bicor")
plotHeatmap(moduleCor_base,
            rowDendro = moduleDendro_base, colDendro = moduleDendro_base,
            file = "comethyl_output/consensus/10_diagnostics/Baseline/Module_Correlation_Heatmap.pdf")

# ── Sample correlation structure ──────────────────────────────────────────────
sampleDendro_base <- getDendro(MEs_base, transpose = TRUE, distance = "bicor")
plotDendro(sampleDendro_base, labelSize = 3, nBreaks = 5,
           file = "comethyl_output/consensus/10_diagnostics/Baseline/Sample_ME_Dendrogram.pdf")

plotHeatmap(MEs_base,
            rowDendro = sampleDendro_base, colDendro = moduleDendro_base,
            legend.title = "Module\nEigennode",
            file = "comethyl_output/consensus/10_diagnostics/Baseline/Sample_ME_Heatmap.pdf")

# Repeat for 36-38wk and Postpartum ...
```

---

## 13. Compute Module Membership (Script 11)

Module membership quantifies how strongly each region correlates with
each module eigengene. High membership in a module (kME close to 1 or
−1) indicates that a region is a core member of that module's
comethylation pattern in a given dataset.

```{r script-11}
# Compute membership for each dataset separately
membership_base <- WGCNA::signedKME(
  datExpr     = t(methAdj_base),
  datME       = MEs_base,
  corFnc      = "bicor",
  corOptions  = "use = 'p'"
)

write.table(membership_base,
            file = "comethyl_output/consensus/11_membership/Baseline/Module_Membership.txt",
            sep = "\t", quote = FALSE, row.names = TRUE)

# Repeat for 36-38wk and Postpartum ...
```

---

## 14. Test ME-Trait Correlations (Script 12a)

`getMEtraitCor()` tests associations between module eigengenes and
sample traits using Bicor or Pearson correlation. Run separately for
each dataset so that dataset-specific trait associations can be compared.

```{r script-12a}
colData_base <- openxlsx::read.xlsx("sample_info_Baseline.xlsx", rowNames = TRUE)

MEtraitCor_base <- getMEtraitCor(
  MEs     = MEs_base,
  colData = colData_base,
  corType = "bicor",
  file    = "comethyl_output/consensus/12a_me_trait_analysis/Baseline/ME_Trait_Correlation_Stats.txt"
)

traitDendro_base <- getCor(MEs_base, y = colData_base,
                           corType = "bicor", robustY = FALSE) %>%
  getDendro(transpose = TRUE)

plotDendro(traitDendro_base, labelSize = 3.5, expandY = c(0.65, 0.08),
           file = "comethyl_output/consensus/12a_me_trait_analysis/Baseline/Trait_Dendrogram.pdf")

plotMEtraitCor(MEtraitCor_base,
               moduleOrder = moduleDendro_base$order,
               traitOrder  = traitDendro_base$order,
               file = "comethyl_output/consensus/12a_me_trait_analysis/Baseline/ME_Trait_Correlation_Heatmap.pdf")

# Repeat for 36-38wk and Postpartum ...
```

---

## 15. Publication-Ready Heatmaps (Script 12b)

Script 12b generates focused heatmaps for user-defined trait subsets
and cross-dataset comparison plots (the same trait set shown
side-by-side for each dataset).

```{r script-12b}
# After running Script 12a for all datasets, subset to traits of interest
traits_of_interest <- c("BMI", "Age", "Parity", "GestationalAge")

MEtraitCor_36wk <- read.table(
  "comethyl_output/consensus/12a_me_trait_analysis/36-38wk/ME_Trait_Correlation_Stats.txt",
  header = TRUE, sep = "\t"
)

# Filter to traits of interest and plot
plotMEtraitCor(
  MEtraitCor_base[MEtraitCor_base$Trait %in% traits_of_interest, ],
  moduleOrder = moduleDendro_base$order,
  file = "comethyl_output/consensus/12b_me_trait_presentation/Baseline/ME_Trait_Selected_Heatmap.pdf"
)
```

---

## 16. Annotate Consensus Modules (Script 15)

Module annotations are performed once since module assignments are
shared across datasets. `annotateModule()` adds nearest gene
information and CpG island context using GREAT and `annotatr`.

```{r script-15}
# Load shared region assignments from Script 09
consensus_regions <- read.table(
  "comethyl_output/consensus/09_consensus_modules/shared/Consensus_Region_Assignments.tsv",
  header = TRUE, sep = "\t"
)

# Annotate all modules
regionsAnno <- annotateModule(
  regions = consensus_regions,
  module  = unique(consensus_regions$Module[consensus_regions$Module != "grey"]),
  genome  = "hg38",
  file    = "comethyl_output/consensus/15_annotation/Annotated_Module_Regions.txt"
)

# Extract gene lists for modules of interest
geneList_turquoise <- getGeneList(regionsAnno, module = "turquoise")
head(geneList_turquoise)
```

---

## From Interactive Exploration to Script-Based Production

The code above is written for interactive learning — running each step
in an R session lets you examine outputs at each stage and tune
parameters before committing to a full run. Once you understand the
pipeline and have settled on parameters, the production workflow uses
numbered scripts (00–15) that are called from SLURM or the command line:

```bash
# Example: run Script 00 for the Baseline dataset
Rscript scripts/consensus/00_import_cpg_reports_consensus.R \
  --project_root /path/to/project \
  --dataset_label Baseline \
  --meta_file /path/to/sample_info.xlsx \
  --cyto_dir /path/to/cytosine_reports/Baseline \
  --workers 8
```

Each script accepts `--help` to list all available arguments. Parameters
that require tuning (coverage thresholds, SD cutoffs, soft power, module
detection settings) are exposed as command-line arguments so you can
document your choices in the run command itself and reproduce the
analysis exactly.

> **Tip.** Set your parameter choices interactively first (as shown in
> this vignette), then move to script-based execution for the final run.
> The script output directories include `run_parameters.txt` and
> `sessionInfo.txt` for every step, providing a complete audit trail.

---

## See Also

- [Function reference](https://lasallegrp.github.io/comethyl/reference/index.html)
- [Get Started vignette](https://lasallegrp.github.io/comethyl/articles/comethyl.html)
- [CpG Cluster Analysis vignette](https://lasallegrp.github.io/comethyl/articles/CpG_Cluster_Analysis.html)
- [Gene Body Analysis vignette](https://lasallegrp.github.io/comethyl/articles/Gene_Body_Analysis.html)
- [Module Preservation vignette](https://lasallegrp.github.io/comethyl/articles/Module_Preservation.html)
